---
title: "Watch Landscape"
subtitle: "A map of 1,000 watches built entirely from their physical specs and prices."
format:
  html:
    page-layout: full
    toc: true
---

```{r setup, include=FALSE}
library(tidyverse)
library(scales)
source("theme_watch.R")

watch_data <- read_csv("1000_Watch_Dataset.csv")

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.align = "center", dev = "png", dpi = 150,
                      dev.args = list(bg = "transparent"))
```

```{r prep, include=FALSE}
# Parse string columns and drop the one non-watch entry (Atmos clock, diameter = 0)
landscape_df <- watch_data %>%
  mutate(
    weight_g    = parse_number(`Weight (grams)`),
    power_res_h = parse_number(`Power Reserve (hours)`)
  ) %>%
  filter(`Case Diameter (mm)` > 0, !is.na(`Price (New)`)) %>%
  drop_na(`Case Diameter (mm)`, `Case Thickness (mm)`, `Lug Width (mm)`,
          `Water Resistance (m)`, weight_g, power_res_h)

feat_names <- c("Case Diameter (mm)", "Case Thickness (mm)", "Lug Width (mm)",
                "Water Resistance (m)", "Price (New)", "weight_g", "power_res_h")

pca_mat <- landscape_df %>% select(all_of(feat_names)) %>% scale()

set.seed(42)
pca_result <- prcomp(pca_mat, center = FALSE, scale. = FALSE)
km_result  <- kmeans(pca_mat, centers = 5, nstart = 25)
var_exp    <- summary(pca_result)$importance[2, 1:2] * 100

scores_df <- landscape_df %>%
  mutate(
    PC1     = pca_result$x[, 1],
    PC2     = pca_result$x[, 2],
    cluster = factor(km_result$cluster),
    broad   = case_when(
      Type %in% c("Smartwatch", "Digital") ~ "Smart / Digital",
      Type == "Luxury"                     ~ "Luxury",
      Type %in% c("Diver", "Tool")         ~ "Diver / Tool",
      Type %in% c("Dress", "Pilot")        ~ "Dress / Pilot",
      TRUE                                 ~ "Sport / Field / GMT"
    )
  )

# --- Give clusters meaningful names, robust to k-means numbering ---
prof <- scores_df %>%
  group_by(cluster) %>%
  summarise(price = median(`Price (New)`),
            wr    = median(`Water Resistance (m)`),
            lug   = median(`Lug Width (mm)`),
            .groups = "drop")

lab   <- setNames(rep(NA_character_, nrow(prof)), as.character(prof$cluster))
avail <- as.character(prof$cluster)
take  <- function(col, fn) {
  sub <- prof[as.character(prof$cluster) %in% avail, ]
  as.character(sub$cluster[fn(sub[[col]])])
}
c_smart <- take("lug",   which.min); lab[c_smart] <- "Smart & Digital";          avail <- setdiff(avail, c_smart)
c_grail <- take("price", which.max); lab[c_grail] <- "Grail & Haute Horlogerie"; avail <- setdiff(avail, c_grail)
c_dive  <- take("wr",    which.max); lab[c_dive]  <- "Dive & Tool";              avail <- setdiff(avail, c_dive)
c_prem  <- take("price", which.max); lab[c_prem]  <- "Premium & Sport";          avail <- setdiff(avail, c_prem)
lab[avail] <- "Dress & Everyday"

label_levels <- c("Dress & Everyday", "Premium & Sport", "Dive & Tool",
                  "Grail & Haute Horlogerie", "Smart & Digital")
scores_df <- scores_df %>%
  mutate(cluster_label = factor(lab[as.character(cluster)], levels = label_levels))

# Biplot loadings scaled to sit inside the scatter
arrow_scale <- max(abs(c(scores_df$PC1, scores_df$PC2))) * 0.72
short_names <- c("Diameter", "Thickness", "Lug Width",
                 "Water Res.", "Price", "Weight", "Power Res.")
loadings_df <- as_tibble(pca_result$rotation[, 1:2]) %>%
  mutate(feature = short_names,
         xend = PC1 * arrow_scale, yend = PC2 * arrow_scale,
         xlab = xend * 1.17,       ylab = yend * 1.17)
```

Every watch in this dataset has at least seven measurable numbers: diameter, thickness, lug width, water resistance, price, weight, and power reserve. Principal component analysis collapses those seven dimensions down to two, preserving as much of the original variance as possible, so each watch becomes a single dot on a map. The further apart two dots sit, the more different those two watches are across their specs.

## What the two axes represent

The arrows below point in the direction each feature pulls most strongly. Arrows pointing the same way mean those features tend to move together. Arrows pointing opposite ways mean they trade off against each other.

```{r biplot, fig.width=9, fig.height=6.5}
ggplot(scores_df, aes(PC1, PC2)) +
  geom_point(aes(colour = broad), alpha = 0.4, size = 1.5) +
  geom_segment(data = loadings_df,
               aes(x = 0, y = 0, xend = xend, yend = yend),
               arrow = arrow(length = unit(0.22, "cm"), type = "closed"),
               colour = "#f0f6fc", linewidth = 0.85) +
  geom_text(data = loadings_df, aes(x = xlab, y = ylab, label = feature),
            size = 3.3, colour = "#f0f6fc", fontface = "bold") +
  scale_colour_watch(name = NULL) +
  labs(
    title = "PCA biplot: seven spec features across the collection",
    subtitle = sprintf("PC1 explains %.1f%% of variance, PC2 %.1f%%", var_exp[1], var_exp[2]),
    x = sprintf("PC1 (%.1f%%)", var_exp[1]),
    y = sprintf("PC2 (%.1f%%)", var_exp[2])
  ) +
  theme_watch() +
  theme(legend.position = "top", legend.text = element_text(size = 10))
```

Price and power reserve typically pull in different directions from water resistance and thickness, which captures the familiar tension in this hobby: the most expensive watches are often the thinnest and least waterproof (dress and luxury), while robust tool watches trade price for depth rating and case volume. Smart and digital watches sit apart because zero lug width is a highly distinctive feature on this dataset.

## The natural families

K-means finds five groupings using all seven features, not just the two PCA axes shown here. Instead of leaving them numbered, each cluster is named for the kind of watch that defines it, based on its own median price, water resistance, and proportions.

```{r clusters, fig.width=9, fig.height=6.5}
ggplot(scores_df, aes(PC1, PC2, colour = cluster_label)) +
  geom_point(alpha = 0.6, size = 1.8) +
  stat_ellipse(aes(group = cluster_label), level = 0.75,
               linewidth = 0.5, linetype = "dashed") +
  scale_colour_watch(name = "Watch family") +
  labs(
    title = "Five watch families, found by clustering on specs",
    subtitle = "Ellipses mark the 75% density region of each family",
    x = sprintf("PC1 (%.1f%%)", var_exp[1]),
    y = sprintf("PC2 (%.1f%%)", var_exp[2])
  ) +
  theme_watch() +
  theme(legend.position = "top")
```

## What defines each family

The heatmap below shows the average of each scaled feature per family. A warm cell means that family scores above average on that feature; a cool cell means below average. This is what justifies each name.

```{r cluster-profiles, fig.width=9, fig.height=5}
profile_long <- scores_df %>%
  select(cluster_label,
         Diameter     = `Case Diameter (mm)`,
         Thickness    = `Case Thickness (mm)`,
         `Lug Width`  = `Lug Width (mm)`,
         `Water Res.` = `Water Resistance (m)`,
         Price        = `Price (New)`,
         Weight       = weight_g,
         `Power Res.` = power_res_h) %>%
  group_by(cluster_label) %>%
  summarise(across(everything(), mean), .groups = "drop") %>%
  pivot_longer(-cluster_label, names_to = "feature", values_to = "mean_val") %>%
  group_by(feature) %>%
  mutate(z = (mean_val - mean(mean_val)) / sd(mean_val)) %>%
  ungroup()

ggplot(profile_long, aes(feature, fct_rev(cluster_label), fill = z)) +
  geom_tile(colour = "#0e1117", linewidth = 1.2) +
  geom_text(aes(label = round(z, 1)), size = 3.4, fontface = "bold",
            colour = ifelse(abs(profile_long$z) > 0.9, "#0e1117", "#c9d1d9")) +
  scale_fill_gradient2(low = "#4CC9F0", mid = "#30363d", high = watch_accent,
                       midpoint = 0, limits = c(-2.2, 2.2), oob = squish,
                       name = "Z-score") +
  labs(
    title = "Family profiles (z-scored feature means)",
    subtitle = "Yellow = above average for that feature, blue = below average",
    x = NULL, y = NULL
  ) +
  theme_watch() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(face = "bold"),
        axis.text.y = element_text(face = "bold", colour = watch_text))
```

Each family now reads on sight. Dive and Tool runs hot on water resistance and case size. Grail and Haute Horlogerie is off the charts on price while staying modest in dimensions. Smart and Digital sits apart on lug width, since every smartwatch reads as zero there. Premium and Sport carries moderate prices with beefier builds, and Dress and Everyday anchors the affordable, slim, low-water-resistance corner.

A few watches will always sit between families, because a Rolex GMT-Master is simultaneously a tool watch and a luxury piece. PCA and k-means do not resolve that ambiguity. What they do is show that the watch market has more distinct shape to it than the usual brand-tier story suggests, and that the most important dividing lines run along specs rather than names.
